---
title: "midterm_report"
author: "Joe LaRocca, Brooklynn McNeil, Malika Top"
date: "2025-03-31"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ISLR)
library(glmnet)
library(caret)
library(pls)
library(tidymodels)
library(corrplot)
library(ggplot2)
library(plotmo)
library(ggrepel)
library(dplyr)
library(pdp)
library(patchwork)
```


# 1. Exploratory Analysis
The aim of this project is to build a prediction model of antibody levels, seeing which demographic and clinical factors influence antibody responses and how antibody levels change over time following vaccination.


#### Loading and splitting data
```{r}
load("./dat1.RData")
load("./dat2.RData")
# split dat1 into a train and test dataset
dat1split = initial_split(data = dat1, prop = 0.8) 
train_dat1 = training(dat1split)
test_dat1 = testing(dat1split)

train_dat1 = train_dat1 |> 
  dplyr::select(-id) |> 
  mutate(gender = factor(gender, levels = c(0, 1), labels = c("female", "male")),
         race = factor(race, levels = c(1, 2, 3, 4), labels = c("white", "asian", 
                                                                "black", "hispanic")),
         smoking = factor(smoking, levels = c(0, 1, 2), labels = c("never smoked",
                                                                   "former smoker",
                                                                   "current smoker")),
         diabetes = factor(diabetes, levels = c(0, 1), labels = c("no", "yes")),
         hypertension = factor(hypertension, levels = c(0, 1), labels = c("no", "yes")))
test_dat1 = test_dat1 |> 
  dplyr::select(-id) |> 
  mutate(gender = factor(gender, levels = c(0, 1), labels = c("female", "male")),
         race = factor(race, levels = c(1, 2, 3, 4), labels = c("white", "asian", 
                                                                "black", "hispanic")),
         smoking = factor(smoking, levels = c(0, 1, 2), labels = c("never smoked",
                                                                   "former smoker",
                                                                   "current smoker")),
         diabetes = factor(diabetes, levels = c(0, 1), labels = c("no", "yes")),
         hypertension = factor(hypertension, levels = c(0, 1), labels = c("no", "yes")))

# keep dat2 untouched for robustness/generalizability purposes later

```
#### Response variable distribution
```{r}
train_dat1 |>
  ggplot(aes(x = log_antibody)) +
  geom_histogram() +
  ggtitle("Distribution of log-transformed antibody levels") +
  labs(tag = "Fig. 1", caption = "The distribution of the response variable is normal so no transformations need to be applied." ) 
```
#### Predictor variable distributions

There are 12 predictor variables in total, 5 that are categorical (gender, race, smoking, diabetes, and hypertension) and 7 that are continuous (age, height, weight, BMI, SBP, LDL). For the categorical predictors, we can visualize the distributions using bar charts, and for continuous predictors using histograms.

```{r}
# gender, race, smoking, diabetes, and hypertension
cat_pred = c("gender", "race", "smoking", "diabetes", "hypertension")
gender_bar = train_dat1 |> 
  ggplot(aes(x = gender, fill = gender)) +
  geom_bar() +
  labs(title = "Distribution of gender")
```


```{r}
#age, height, weight, BMI, SBP, LDL
cont_pred = c("age", "height", "weight", "BMI", "SBP", "LDL")
age_hist = train_dat1 |> 
  ggplot(aes(x = age)) +
  geom_histogram() +
  labs(title = "Distribution of age")
height_hist = train_dat1 |> 
  ggplot(aes(x = height)) +
  geom_histogram() +
  labs(title = "Distribution of height")
weight_hist = train_dat1 |> 
  ggplot(aes(x = weight)) +
  geom_histogram() +
  labs(title = "Distribution of weight")
bmi_hist = train_dat1 |> 
  ggplot(aes(x = bmi)) +
  geom_histogram() +
  labs(title = "Distribution of BMI")
sbp_hist = train_dat1 |> 
  ggplot(aes(x = SBP)) +
  geom_histogram() +
  labs(title = "Distribution of SBP")
ldl_hist = train_dat1 |> 
  ggplot(aes(x = LDL)) +
  geom_histogram() +
  labs(title = "Distribution of LDL")
```


#### Visualizing relationship between predictor-response variables
For visualizing the relationship between the predictor and response variables, we can make boxplots for the categorical variables, and scatterplots for the continuous. 
```{r}
gender_box = train_dat1 |> 
  ggplot(aes(x = gender, y = log_antibody, fill = gender)) +
  geom_boxplot() +
  labs(title = "Distribution of log-antibody levels by gender")
race_box = train_dat1 |> 
  ggplot(aes(x = race, y = log_antibody, fill = race)) +
  geom_boxplot() +
  labs(title = "Distribution of log-antibody levels by race")
smoking_box = train_dat1 |> 
  ggplot(aes(x = smoking, y = log_antibody, fill = smoking)) +
  geom_boxplot() +
  labs(title = "Distribution of log-antibody levels by smoking status")
diabetes_box = train_dat1 |> 
  ggplot(aes(x = diabetes, y = log_antibody, fill = diabetes)) +
  geom_boxplot() +
  labs(title = "Distribution of log-antibody levels by diabetes status")
hypertension_box = train_dat1 |> 
  ggplot(aes(x = hypertension, y = log_antibody, fill = hypertension)) +
  geom_boxplot() +
  labs(title = "Distribution of log-antibody levels by hypertension status")
(gender_box | race_box | smoking_box) /
      (diabetes_box | hypertension_box)
```

```{r}
cont_pred = c("age", "height", "weight", "bmi", "SBP", "LDL", "time")
scatter_plots = list()
for (pred in cont_pred) {
  plot = train_dat1 |> 
  ggplot(aes(x = !!sym(pred), y = log_antibody)) +
  geom_point() +
  geom_smooth() 
  scatter_plots[[pred]] = plot
}
wrap_plots(scatter_plots)
age_scatter = train_dat1 |> 
  ggplot(aes(x = age, y = log_antibody)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship of age and log-antibody levels") 
height_scatter = train_dat1 |> 
  ggplot(aes(x = height, y = log_antibody)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship of height and log-antibody levels") 
weight_scatter = train_dat1 |> 
  ggplot(aes(x = weight, y = log_antibody)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship of weight and log-antibody levels") 
bmi_scatter = train_dat1 |> 
  ggplot(aes(x = bmi, y = log_antibody)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship of BMI and log-antibody levels") 
sbp_scatter = train_dat1 |> 
  ggplot(aes(x = sbp, y = log_antibody)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship of SBP and log-antibody levels") 
ldl_scatter = train_dat1 |> 
  ggplot(aes(x = ldl, y = log_antibody)) +
  geom_point() +
  geom_smooth() +
  labs(title = "Relationship of LDL and log-antibody levels") 
```
There appear to be some non-linear relationships between BMI, time, and the log-antibody levels.  

#### Checking for correlation between predictors
```{r}
x = model.matrix(log_antibody ~ ., train_dat1)[,-1]
y = train_dat1$log_antibody
```

# 2. Model training
```{r}

```

# 3. Results
















